Design and Implementation of Probabilistic Programming Language Anglican

### 2016-08-04 paper/paper.tex
update

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values
%Clojure objects
suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm} and \texttt{fm} (Anglican counterparts of Clojure
\texttt{defn} and \texttt{fn}). Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe} (as well \texttt{predict}
for the output), and can be seamlessly called from inside
Anglican queries, just like functions locally defined within the
same query.

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code, however in Clojure, unlike in
Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Within \texttt{defquery},
\texttt{let}, \texttt{if}, \texttt{when}, \texttt{cond},
\texttt{case}, \texttt{and}, \texttt{or}, \texttt{fn} forms are
supported (others may be in the future but are not now). 
In \texttt{let} bindings and \texttt{fn} argument lists,
vector destructuring (but not hash map destructuring) is
supported. Compound literals for vectors, hash maps, and
sets are supported just like in Clojure.

\subsection{\texttt{recur}}

Anglican is stackless, therefore \texttt{recur} is
unnecessary, no recursive call can lead to stack overflow;
Recursive calls to functions should be used instead. However,
\texttt{loop}/\texttt{recur} is provided for convenience as a way to express
loops. \texttt{recur} outside of \texttt{loop} will lead to unpredictable
behaviour and hard-to-catch errors.

@@

All of Clojure's core library except for higher-order functions
(functions that accept other functions as arguments) is
available in Anglican. Higher-order functions cannot be reused
from Clojure, they have to be re-implemented to accept
functional arguments in CPS form. The following higher-order
functions are implemented: \texttt{map}, \texttt{reduce},
\texttt{filter}, \texttt{some}, \texttt{repeatedly},
\texttt{comp}, \texttt{partial}.

>>>>

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values
%Clojure objects
suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm} 
%and \texttt{fm} 
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}). 
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe} (as well \texttt{predict}
for the output), and can be seamlessly called from inside
Anglican queries, just like functions locally defined within the
same query.

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code, however in Clojure, unlike in
Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Anglican queries
are defined within \texttt{defquery} as shown in the previous section,
using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case}, 
\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
other forms of Clojure may be in the future but are not now. 
In \texttt{let} bindings and \texttt{fn} argument lists,
Clojure's pattern matching mechanism for vector data type
(called vector destructuring) is supported. Also, compound 
literals for vectors, hash maps, and
sets are supported just like in Clojure.

\subsection{Tail recursion}

Clojure provides special constructs \texttt{loop} and \texttt{recur}
for writing tail-recursive programs. Such constructs are not necessary
in Anglican, because Anglican programs are CPS-converted and do not
use stacks; no recursive call in Anglican can lead to stack overflow.
In Anglican, recursive calls to functions should be used instead of
\texttt{recur}. However, \texttt{loop}/\texttt{recur} is provided in Anglican
for convenience as a way to express loops. \texttt{recur} outside of \texttt{loop} 
will lead to unpredictable behaviour and hard-to-catch errors.

@@

All of Clojure's core library except for higher-order functions
(functions that accept other functions as arguments) is
available in Anglican. Higher-order functions cannot be reused
from Clojure, as they have to be re-implemented to accept
functional arguments in CPS form. The following higher-order
functions are implemented: \texttt{map}, \texttt{reduce},
\texttt{filter}, \texttt{some}, \texttt{repeatedly},
\texttt{comp}, \texttt{partial}.
